Quotient maps and configuration spaces of hard disks

### 2022-02-09 manuscript/old/main_010222.tex
manusctript

Configuration spaces of hard disks have been studied previously \cite{carlsson2012harddisks,baryshnikov2014min}. Ritchey \cite{ritcheyphd} specifically studied the configuration spaces of hard disks on the hexagonal torus. \red{They provided appropriate definitions of critical points in the configuration space of hard disks and the associated critical index in the context of Morse theory (a detailed explanation is provided in Secs. \ref{sec:tautological_function} and \ref{sec:morse_theory}), and examined the equivalence classes that critical points form under the action of translation, permutation and discrete lattice symmetries.} A high density of critical points around the packing fraction of the solid-liquid transition indicated rapidly-changing configuration space topology there. This is suggestive of idea underlying the Topological Hypothesis, i.e., that a signature of two-dimensional hard disk melting should be visible in the distribution of critical points of the potential energy surface in the corresponding configuration space. One area not comprehensively addressed by this earlier work is the effect that quotienting out by the action of symmetry groups has on the number and distribution of the critical points.

@@

The tautological function $\tau: \Lambda(n) \rightarrow R$ is defined as
\begin{equation*}
\tau = \min\limits_{\substack{1 \leq i < j \leq n}} {r_{ij}}
\label{eq:tautological_function}
\end{equation*}
where $r_{ij}$ is half the geodesic distance between the centers of disks $i$ and $j$. Intuitively, $\tau$ is the maximum radius that the disks could have without any pair overlapping given the positions of the disk centers. Observe that the configuration space
\begin{equation}
\Gamma(n,\rho) = \tau^{-1}\left([\rho,\infty)\right)
\label{eq:configuration_space2}
\end{equation}
of $n$ hard disks of radius $\rho$ is the superlevel set of $\tau$, or the set of all configurations that could accommodate disks of radius at least $\rho$.

@@

\red{At this point, it is worth mentioning that the critical points considered in this work are not the critical points of the potential energy surface on the configuration space $\Gamma(n,\rho)$, but the critical points of the tautological function $\tau$ on $\Lambda(n)$. Observe that $\Gamma(n,\rho)$ is some particular subset of $\Lambda(n)$ containing the points in $\Lambda(n)$ such that no overlapping will occur if disks of radius $\rho$ are placed to these points. In the context of Morse theory, the critical points of $\tau$ are then the locations where the topology of $\Lambda(n)$ (hence $\Gamma(n,\rho)$) changes. Keeping these in mind, the potential energy of each point in $\Gamma(n,\rho)$ is necessarily zero since they are guaranteed to be non-overlapping.}

>>>>

Configuration spaces of hard disks have been studied previously \cite{carlsson2012harddisks,baryshnikov2014min}. Ritchey \cite{ritcheyphd} specifically studied the configuration spaces of hard disks on the hexagonal torus. They provided appropriate definitions of critical points and critical index, and examined the equivalence classes that critical points form under the action of translation, permutation and discrete lattice symmetries. A high density of critical points around the packing fraction of the solid-liquid transition indicated rapidly-changing configuration space topology there. This is suggestive of idea underlying the Topological Hypothesis, i.e., that a signature of two-dimensional hard disk melting should be visible in the distribution of critical points of the potential energy surface in the corresponding configuration space. One area not comprehensively addressed by this earlier work is the effect that quotienting out by the action of symmetry groups has on the number and distribution of the critical points.

@@

The tautological function $\tau: \Lambda(n) \rightarrow R$ is defined as
\begin{equation*}
\tau = \min\limits_{\substack{1 \leq i < j \leq n}} {r_{ij}}
\label{eq:tautological_function}
\end{equation*}
where $r_{ij}$ is half the geodesic distance between the centers of disks $i$ and $j$. Intuitively, $\tau$ is the maximum radius that the disks could have without any pair overlapping given the positions of the disk centers. Observe that the configuration space
\begin{equation}
\Gamma(n,\rho) = \tau^{-1}[\rho,\infty)
\label{eq:configuration_space2}
\end{equation}
of $n$ hard disks of radius $\rho$ is the superlevel set of $\tau$, or the set of all configurations that could accommodate disks of radius at least $\rho$.

@@
